---
title: "Demography and Epidemic-Demography Coupling"
format: html
---

Demography matters because age structure changes over time. Births, mortality,
ageing, and migration all reshape the population seen by an epidemic model, so
the same contact matrix can produce different trajectories as the age mix
shifts.

This tutorial keeps the mechanics visible in two stages:

1. run a demographic-only workflow with `simulate_demography()`;
2. use the same kind of demographic process inside `simulate_deterministic()`
   for a coupled epidemic-demography run.

For deeper background, see [docs/demography_workflows.md](../docs/demography_workflows.md),
[docs/demographic_residuals.md](../docs/demographic_residuals.md),
[docs/model_conventions.md](../docs/model_conventions.md),
[examples/mock_demographic_workflow.R](../examples/mock_demographic_workflow.R),
[examples/mock_seir_demography.R](../examples/mock_seir_demography.R), and
[examples/annual_cohort_sir_demography.R](../examples/annual_cohort_sir_demography.R).

```{r}
if ("package:agepi" %in% search()) {
  # Package already available.
} else if (dir.exists("R")) {
  invisible(lapply(list.files("R", pattern = "[.]R$", full.names = TRUE), source))
} else if (requireNamespace("agepi", quietly = TRUE)) {
  library(agepi)
} else {
  stop(
    "Package agepi is not available. Run this tutorial from the package root ",
    "or install agepi first.",
    call. = FALSE
  )
}

age_structure <- AgeStructure(
  age_groups = c("0-4", "5-9", "10+"),
  lower_bounds = c(0, 5, 10),
  upper_bounds = c(4, 9, Inf)
)

fertility <- FertilitySchedule(
  data.frame(
    time = c(0, 1),
    age_group = c("10+", "10+"),
    fertility_rate = c(0.025, 0.024),
    stringsAsFactors = FALSE
  ),
  age_structure
)

mortality <- MortalitySchedule(
  data.frame(
    time = rep(c(0, 1), each = age_structure$n_age_groups),
    age_group = rep(age_structure$age_groups, times = 2),
    mortality_rate = c(0.004, 0.002, 0.015, 0.004, 0.002, 0.016),
    stringsAsFactors = FALSE
  ),
  age_structure
)

migration <- MigrationSchedule(
  data.frame(
    time = rep(c(0, 1), each = age_structure$n_age_groups),
    age_group = rep(age_structure$age_groups, times = 2),
    migration_count = c(2, -1, 1, 2, -1, 1),
    stringsAsFactors = FALSE
  ),
  age_structure
)

process <- build_demographic_process(
  age_structure = age_structure,
  fertility_schedule = fertility,
  mortality_schedule = mortality,
  migration_schedule = migration,
  mode = "migration"
)

initial_population <- c(500, 600, 900)

demography_only <- simulate_demography(
  process = process,
  initial_state = initial_population,
  times = c(0, 1, 2),
  time_policy = "step"
)

head(demography_only, 9)

contact_matrix <- matrix(
  c(
    4, 2, 1,
    2, 5, 2,
    1, 2, 4
  ),
  nrow = age_structure$n_age_groups,
  byrow = TRUE
)

epidemic_state <- data.frame(
  compartment = rep(c("S", "E", "I", "R"), each = age_structure$n_age_groups),
  age_group = rep(age_structure$age_groups, times = 4),
  value = c(
    c(497, 597, 897),
    c(3, 2, 1),
    c(2, 1, 1),
    c(0, 0, 0)
  ),
  stringsAsFactors = FALSE
)

coupled <- simulate_deterministic(
  initial_state = epidemic_state,
  times = seq(0, 1, by = 0.25),
  model = SEIRModel(sigma = 0.4, gamma = 0.25),
  age_structure = age_structure,
  contact_matrix = contact_matrix,
  beta = 0.08,
  demographic_process = process,
  time_policy = "step",
  method = "euler"
)

head(coupled, 12)
compartment_totals(coupled)
total_population(coupled)
```

The first run is demographic-only: it returns a population table with `time`,
`age_group`, and `population`. That is the right choice when you want to check
fertility, mortality, migration, or ageing in isolation.

The second run couples the same style of demographic process to an epidemic
model. Here the demographic schedules change the population seen by the disease
model, while the disease compartments still evolve through their own
transitions.

The annual-cohort pathway is an advanced option for one-year age grids. It is
useful when you want cohort-style demographic stepping, but it is not the
default learning path.

```{r}
if (requireNamespace("deSolve", quietly = TRUE)) {
  ages_1yr <- wpp_age_structure_1year(max_age = 5)

  fertility_1yr <- FertilitySchedule(
    data.frame(
      time = 2020,
      age_group = "3",
      fertility_rate = 0.08,
      stringsAsFactors = FALSE
    ),
    ages_1yr
  )

  mortality_1yr <- MortalitySchedule(
    data.frame(
      time = 2020,
      age_group = ages_1yr$age_groups,
      mortality_rate = c(0.01, 0.002, 0.001, 0.001, 0.002, 0.03),
      stringsAsFactors = FALSE
    ),
    ages_1yr
  )

  annual_process <- build_demographic_process(
    age_structure = ages_1yr,
    fertility_schedule = fertility_1yr,
    mortality_schedule = mortality_1yr
  )

  annual_state <- data.frame(
    compartment = rep(c("S", "I", "R"), each = ages_1yr$n_age_groups),
    age_group = rep(ages_1yr$age_groups, times = 3),
    value = c(
      c(998, 947, 896, 846, 797, 1195),
      c(2, 3, 4, 4, 3, 5),
      c(0, 0, 0, 0, 0, 0)
    ),
    stringsAsFactors = FALSE
  )

  annual_contacts <- matrix(1, nrow = ages_1yr$n_age_groups, ncol = ages_1yr$n_age_groups)
  diag(annual_contacts) <- 3

  simulate_deterministic(
    initial_state = annual_state,
    times = c(2020, 2021),
    model = SIRModel(gamma = 0.25),
    age_structure = ages_1yr,
    contact_matrix = annual_contacts,
    beta = 0.03,
    demographic_process = annual_process,
    ageing_policy = "annual_cohort",
    time_policy = "step"
  )
} else {
  message("Annual-cohort coupling is optional here and needs deSolve.")
}
```

The key distinction is simple:

- demographic-only workflows update population structure without disease;
- coupled workflows let the epidemic model and the demographic process evolve
  together.

That separation keeps the model readable and makes it easier to decide when a
population change belongs to the demographic layer and when it belongs to the
disease layer.
