---
title: "Stochastic Simulation"
format: html
---

This tutorial introduces fixed-population stochastic simulation in `agepi` and
shows how it sits alongside the deterministic workflows from the earlier
tutorials.

It is a research prototype, not a calibration or forecasting workflow. The
current stochastic simulator is fixed-population only: it does not yet include
stochastic demography, ageing, fertility, mortality, or migration.

For more detail, see [docs/stochastic_simulation_design.md](../docs/stochastic_simulation_design.md),
[docs/epidemic_model_workflows.md](../docs/epidemic_model_workflows.md),
[examples/stochastic_sir.R](../examples/stochastic_sir.R),
[examples/stochastic_seir.R](../examples/stochastic_seir.R), and
[examples/stochastic_cumulative_flows.R](../examples/stochastic_cumulative_flows.R).

```{r}
if ("package:agepi" %in% search()) {
  # Package already available.
} else if (dir.exists("R")) {
  invisible(lapply(list.files("R", pattern = "[.]R$", full.names = TRUE), source))
} else if (requireNamespace("agepi", quietly = TRUE)) {
  library(agepi)
} else {
  stop(
    "Package agepi is not available. Run this tutorial from the package root ",
    "or install agepi first.",
    call. = FALSE
  )
}

age_structure <- AgeStructure(
  age_groups = c("0-4", "5-9", "10-14"),
  lower_bounds = c(0, 5, 10),
  upper_bounds = c(4, 9, 14)
)

population <- c(200, 250, 180)
initial_infections <- c(3, 2, 1)
initial_state <- data.frame(
  compartment = rep(c("S", "I", "R"), each = age_structure$n_age_groups),
  age_group = rep(age_structure$age_groups, times = 3),
  value = c(population - initial_infections, initial_infections, rep(0, 3)),
  stringsAsFactors = FALSE
)

contact_matrix <- matrix(
  c(
    4, 2, 1,
    2, 5, 2,
    1, 2, 4
  ),
  nrow = age_structure$n_age_groups,
  byrow = TRUE
)

times <- seq(0, 10, by = 1)
model <- SIRModel(gamma = 0.25)

stochastic <- simulate_stochastic(
  initial_state = initial_state,
  times = times,
  model = model,
  age_structure = age_structure,
  contact_matrix = contact_matrix,
  beta = 0.08,
  seed = 123,
  return_events = TRUE
)

deterministic <- simulate_deterministic(
  initial_state = initial_state,
  times = times,
  model = model,
  age_structure = age_structure,
  contact_matrix = contact_matrix,
  beta = 0.08
)

head(stochastic$trajectory, 12)
head(stochastic$events, 10)
head(deterministic, 12)
aggregate(value ~ time + compartment, stochastic$trajectory, sum)
aggregate(value ~ time + compartment, deterministic, sum)
```

The deterministic and stochastic runs use the same core model inputs:

- the same `AgeStructure()`;
- the same initial compartment-by-age state;
- the same recipient-row, source-column contact matrix;
- the same disease model.

The difference is in how transitions happen. Deterministic simulation evolves
the expected trajectory, while stochastic simulation samples realised events.
That makes stochastic output useful when you want sample paths, event timing, or
variability around a deterministic baseline.

The return shape is still familiar. The stochastic simulator returns the usual
`trajectory` table, and `return_events = TRUE` adds a realised event log with
event times and transition identifiers. That log is often the best place to
inspect what happened on a particular sample path.

```{r}
head(stochastic$events, 6)[, c("time", "event", "transition_id", "age_group", "from", "to")]
```

Because the event log is explicit, it is easy to derive simple diagnostics:

```{r}
stochastic_event_counts <- aggregate(
  time ~ transition_id + age_group,
  stochastic$events,
  length
)
names(stochastic_event_counts)[names(stochastic_event_counts) == "time"] <- "realised_events"
stochastic_event_counts
```

If you need cumulative flow counts, supply `cumulative_flows` and inspect the
returned `cumulative` table. Those counts are derived from realised events and
stay separate from the disease compartments themselves.

```{r}
stochastic_cumulative <- simulate_stochastic(
  initial_state = initial_state,
  times = times,
  model = model,
  age_structure = age_structure,
  contact_matrix = contact_matrix,
  beta = 0.08,
  seed = 123,
  return_events = TRUE,
  cumulative_flows = list(
    infections = list(from = "S", to = "I"),
    recoveries = list(from = "I", to = "R")
  )
)

head(stochastic_cumulative$cumulative, 12)
```

The main practical checks are straightforward:

- use `seed` when you want reproducible sample paths;
- inspect `trajectory` when you want the realised compartment time series;
- inspect `events` when you want the event-level history;
- compare against deterministic output when you want a baseline shape check.

For the package boundary, the important point is simple: stochastic simulation
currently covers fixed-population epidemic events only. If your question depends
on stochastic births, deaths, ageing, or migration, that is outside the current
stochastic surface.
