---
title: "Demography-Only Workflows and Diagnostics"
format: html
---

This tutorial gives standalone demography its own guided path, separate from
epidemic-demography coupling.

It is useful when you want to check age-specific population change on its own,
compare a demographic process against observed or reference values, or prepare
the population layer that a later epidemic workflow will consume.

For more detail, see [docs/demography_workflows.md](../docs/demography_workflows.md),
[docs/demographic_residuals.md](../docs/demographic_residuals.md),
[examples/mock_demographic_workflow.R](../examples/mock_demographic_workflow.R),
[examples/demography_plots.R](../examples/demography_plots.R), and
[examples/wpp_demography_validation.R](../examples/wpp_demography_validation.R).

```{r}
if ("package:agepi" %in% search()) {
  # Package already available.
} else if (dir.exists("R")) {
  invisible(lapply(list.files("R", pattern = "[.]R$", full.names = TRUE), source))
} else if (requireNamespace("agepi", quietly = TRUE)) {
  library(agepi)
} else {
  stop(
    "Package agepi is not available. Run this tutorial from the package root ",
    "or install agepi first.",
    call. = FALSE
  )
}

age_structure <- AgeStructure(
  age_groups = c("0-4", "5-9", "10+"),
  lower_bounds = c(0, 5, 10),
  upper_bounds = c(4, 9, Inf)
)

fertility <- FertilitySchedule(
  data.frame(
    time = c(0, 1, 2),
    age_group = c("10+", "10+", "10+"),
    fertility_rate = c(0.02, 0.02, 0.02),
    stringsAsFactors = FALSE
  ),
  age_structure
)

mortality <- MortalitySchedule(
  data.frame(
    time = rep(c(0, 1, 2), each = age_structure$n_age_groups),
    age_group = rep(age_structure$age_groups, times = 3),
    mortality_rate = c(
      0.01, 0.005, 0.02,
      0.01, 0.005, 0.02,
      0.01, 0.005, 0.02
    ),
    stringsAsFactors = FALSE
  ),
  age_structure
)

migration <- MigrationSchedule(
  data.frame(
    time = rep(c(0, 1, 2), each = age_structure$n_age_groups),
    age_group = rep(age_structure$age_groups, times = 3),
    migration_count = c(
      1, -1, 0,
      1, -1, 0,
      1, -1, 0
    ),
    stringsAsFactors = FALSE
  ),
  age_structure
)

process <- build_demographic_process(
  age_structure = age_structure,
  fertility_schedule = fertility,
  mortality_schedule = mortality,
  migration_schedule = migration,
  mode = "migration"
)

initial_population <- c(500, 450, 800)
demography_only <- simulate_demography(
  process = process,
  initial_state = initial_population,
  times = c(0, 1, 2),
  time_policy = "step"
)

head(demography_only, 9)
aggregate(population ~ time, demography_only, sum)
```

Demographic-only workflows are useful because they keep the mechanics visible:

- fertility adds new people to the population process;
- mortality removes people from the active population;
- ageing moves people between age groups;
- migration can add or remove people according to the declared policy.

That makes `simulate_demography()` a good place to check whether your age grid
and schedules behave the way you expect before you add a disease model on top.

The output is a demographic population table with `time`, `age_group`, and
`population` columns. It is intentionally separate from the epidemic trajectory
format used by `simulate_deterministic()`.

If you want a quick diagnostic comparison against observed data, the package
provides residual helpers that stay in the demographic layer:

```{r}
observed <- Demography(
  data.frame(
    time = rep(c(0, 1, 2), each = age_structure$n_age_groups),
    age_group = rep(age_structure$age_groups, times = 3),
    population = c(
      500, 450, 800,
      495, 455, 810,
      490, 460, 820
    ),
    stringsAsFactors = FALSE
  ),
  age_structure
)

comparison <- compare_demography_to_observed(demography_only, observed)
summary <- summarise_demography_comparison(comparison)
residual <- implied_demographic_residual(observed, process)

summary
head(comparison, 6)
head(residual, 6)
```

The diagnostic helpers do not fit the model for you. They show the gap between
the simulated population and the observed or reference trajectory, which is
often enough to see whether fertility, mortality, migration, or ageing needs
attention.

If you have a reporting grid that is broader than the model grid, use the age
grid mapping helpers to aggregate output explicitly rather than assuming the
model grid and reporting grid are interchangeable.

```{r}
reporting <- AgeStructure(
  age_groups = c("0-9", "10+"),
  lower_bounds = c(0, 10),
  upper_bounds = c(9, Inf)
)

mapping <- AgeGridMapping(age_structure, reporting, open_ended = "include")
aggregate_demography_trajectory_age_grid(demography_only, mapping)
```

Demographic-only outputs are also the cleanest starting point for later
epidemic-demography workflows. Once the population layer is checked, the same
age structure and schedule conventions can be reused inside
`simulate_deterministic()` with a disease model attached.

```{r}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plots <- plot_demography(demography_only)
  names(plots)
} else {
  message("plot_demography() is optional here and needs ggplot2.")
}
```

The plotting step is optional. If you only need the mechanics, the tables above
are enough to validate the demographic process.
