---
title: "External Data and Case-Study Scaffolds"
format: html
---

This tutorial explains how `agepi` connects transparent model workflows to
optional external data sources and realistic-but-not-calibrated case-study
scaffolds.

The main idea is a boundary, not a new model family: core modelling inputs stay
in `agepi` objects, while external-data adapters prepare optional source data
for those objects.

For more detail, see [docs/external_data_adapters.md](../docs/external_data_adapters.md),
[docs/contact_matrix_workflows.md](../docs/contact_matrix_workflows.md),
[docs/demography_workflows.md](../docs/demography_workflows.md),
[examples/observed_case_age_groups.R](../examples/observed_case_age_groups.R),
[examples/epiparameter_seir.R](../examples/epiparameter_seir.R),
[examples/wpp_projection_backed_demography.R](../examples/wpp_projection_backed_demography.R),
and [examples/kiribati_tb_realistic_demography.R](../examples/kiribati_tb_realistic_demography.R).

```{r}
if ("package:agepi" %in% search()) {
  # Package already available.
} else if (dir.exists("R")) {
  invisible(lapply(list.files("R", pattern = "[.]R$", full.names = TRUE), source))
} else if (requireNamespace("agepi", quietly = TRUE)) {
  library(agepi)
} else {
  stop(
    "Package agepi is not available. Run this tutorial from the package root ",
    "or install agepi first.",
    call. = FALSE
  )
}
```

Core inputs are the pieces the model actually consumes:

- `AgeStructure()` for the age grid;
- compartment-by-age state tables;
- contact matrices in recipient-row, source-column orientation;
- disease models such as `SIRModel()`, `SEIRModel()`, or `CompartmentModel()`;
- demographic schedules when the model includes population change.

External adapters sit one step upstream. They help with tasks such as mapping
observed cases into the model age grid, translating WPP-style demography into
`agepi` schedules, or loading and adapting a published contact-matrix source.

The smallest example is observed-case age mapping:

```{r}
age_structure <- AgeStructure(
  age_groups = c("0-4", "5-9", "10+"),
  lower_bounds = c(0, 5, 10),
  upper_bounds = c(4, 9, Inf)
)

observed_cases <- data.frame(
  case_id = paste0("case_", 1:6),
  onset_day = c(0, 0, 1, 1, 2, 2),
  age = c(2, 8, 11, 36, 4, 7),
  stringsAsFactors = FALSE
)

agepi_cases <- as_agepi_cases(
  observed_cases,
  age_structure = age_structure,
  age_col = "age"
)

incident_targets <- aggregate(
  case_id ~ onset_day + age_group,
  agepi_cases,
  length
)
names(incident_targets)[names(incident_targets) == "case_id"] <- "cases"

agepi_cases
incident_targets
```

That pattern matters because observed data rarely arrive in exactly the same
age bins as the model. Mapping them explicitly keeps the age convention visible
and makes later calibration or diagnostic work less ambiguous.

Contact matrices follow the same logic. The package separates the source layer
from the model layer so that provenance, adaptation, and orientation stay
clear. In a small toy workflow, a plain numeric matrix may be enough. In a
larger scaffold, the model may load a source matrix and adapt it to the model
age grid.

```{r}
contact_matrix <- matrix(
  c(
    4, 2, 1,
    2, 5, 2,
    1, 2, 4
  ),
  nrow = age_structure$n_age_groups,
  byrow = TRUE
)

contact_matrix
```

The external-data adapter story is broader than any single example:

- WPP-style demographic tables can be standardised into fertility, mortality,
  and migration schedules;
- contact-matrix sources can be loaded and adapted to the simulation age grid;
- epiparameter objects can provide a high-level way to convert a delay
  distribution into an approximate transition rate;
- validation and diagnostic examples can compare simulated output to observed
  or reference data without turning the package into a black-box fitting tool.

Optional rate conversion from `epiparameter` looks like this:

```{r}
if (requireNamespace("epiparameter", quietly = TRUE)) {
  incubation <- epiparameter::epiparameter_db(
    disease = "COVID-19",
    epi_name = "incubation period",
    single_epiparameter = TRUE
  )
  rate_from_epiparameter(incubation)
} else {
  message("epiparameter is optional here; see examples/epiparameter_seir.R.")
}
```

For WPP-style demography, the package keeps the distinction clear between
standardising input tables and simulating the resulting process. The adapter
layer prepares the schedules, while `simulate_demography()` and
`simulate_deterministic()` consume the resulting `DemographicProcess()`.

```{r}
if (requireNamespace("wpp2024", quietly = TRUE)) {
  message("See examples/wpp_projection_backed_demography.R and examples/kiribati_tb_realistic_demography.R for the full WPP-backed pathway.")
} else {
  message("wpp2024 is optional here; the WPP-backed examples are linked for reference.")
}
```

The Kiribati TB scaffold is the best example of how these pieces fit together
in a realistic but still provisional workflow. It combines WPP-backed
demography, an adapted contact matrix, and a TB-style compartment model, but it
remains a scaffold rather than a calibrated policy model.

That caution matters. Case-study scaffolds are useful when you want to:

- show how the package fits together on a real-world problem;
- keep every modelling assumption transparent;
- explore scenario structure before calibration;
- identify which external data are missing.

They are not the same as a fitted or policy-ready model. For Kiribati TB in
particular, the public-data scaffold should be read as a transparent starting
point whose assumptions still need calibration, validation, and local evidence.

The short version is simple: core modelling inputs live inside `agepi`, while
optional adapters and case-study scaffolds help you bring external data into
that framework without hiding the assumptions.
